### 2020-09-06 case-studies.tex
popl21 reviews

\section{Case Studies}\label{sec:case-studies}
Let us begin by considering livelits from the client programmer's perspective by way of
two domain-specific case studies:
a course grade assignment case study in Sec.~\ref{sec:live-grading}
and an image transformation case study in Sec.~\ref{sec:image-transformation}.
We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}.\todo{do we?}{}
These case studies and examples have been implemented
in Hazel, a browser-based live programming environment for a pure typed functional language in the
ML family. We assume basic familiarity with ML.

\subsection{Case Study: Grading with Livelits}\label{sec:live-grading}
\begin{figure}
    \includegraphics[width=30pc]{grade-cutoff-livelit.png}
\caption{Grading with Livelits}
\label{fig:grading}
\end{figure}

@@

Perhaps the most common end-user application in this domain is the spreadsheet.
A spreadsheet allows the instructor to record student grades using a natural tabular interface,
visualize this data in one of a finite number of plot styles, and perform basic computations on this data,
with results updated live, i.e. automatically after each cell has been edited.
However, these affordances are limited. It is difficult to package up common operations and workflows
into reusable libraries, interact with the data using domain-specific visualizations,
and perform operations that have not been anticipated by the designers of the spreadsheet system
(e.g. operations that prepare the data in an idiosyncratic format suitable for automated submission
to the university's grading system).
General-purpose programming languages equipped with suitable libraries for working with numeric data
are flexible enough to handle these scenarios, but users of these languages
sacrifice the ability to receive live feedback and directly manipulate data and visualizations thereof.

Livelits are able to address this tension.
Fig.~\ref{fig:grading} shows a Hazel program where the user has:
\begin{enumerate}
\item recorded numeric \li{grades} for each student using a livelit, \li{\$dataframe}, that implements a tabular user interface with support for live evaluation of the Hazel expression in each cell,
\item programmatically computed the overall \li{averages} for each student by applying a function, \li{compute_averages}, defined in a library  (not shown) shared between multiple courses,
\item determined reasonable \li{cutoffs} between letter grades by directly manipulating a livelit that offers a domain-specific drag-and-drop interface superimposed on a live visualization of the grade distribution
(for the sake of space, we plot dots rather than a full histogram),
\item programmatically assigned letter grades to each student based on the \li{cutoffs} determined in step (3) by calling another shared library function, \li{assign_grades},
\item formatted the final grade assignments in a manner suitable for submission to the university system by calling another shared library function, \li{format_for_university}.
\end{enumerate}

\subsubsection{Livelit Expansion}\label{sec:livelit-expansion}
Livelit expressions are given dynamic meaning by expansion to core expressions, i.e. expressions that do not contain livelits.
For example, the expansion of Fig.~\ref{fig:grading} is:

\begin{lstlisting}
let grades = Dataframe (
  ["A1", "A2", "A3", "A4", "Midterm", "Final"],
  [
    ("Alice", [24. +. 36. +. 33., 92., 83.5, 87.5, 95., 88.]),
    ("Bob", [61., 64., 98., 89.5, 70., 85.]),
    ("Carol", [75., 81., 73., 71., 82., 79.]),
    (* ... *)
  ]) in
let averages = compute_averages grades weights in
let cutoffs = (.A 90.0, .B 80.0, .C 70.0, .D 60.0) in
format_for_university (assign_grades averages cutoffs)
\end{lstlisting}

The programmer can inspect this expansion in Hazel.
Ideally, however, it would not be necessary to inspect the expansion
(or the livelit implementation, which specifies the expansion logic and is discussed in Sec.~\ref{sec:livelit-definitions})
to reason about types and binding.
After all, one does not need to look inside
function bodies to reason about types and binding.
Instead, in the words of \citet{DBLP:conf/ifip/Reynolds83},
``type structure is a syntactic discipline for maintaining levels of abstraction''.
The livelits mechanism supports reasoning abstractly about types and binding by
several means.

To support abstract reasoning about the type of the expansion,
livelit definitions declare an \emph{expansion type}.
The declarations of the two livelits in Fig.~\ref{fig:grading},
eliding their implementations, are:
\begin{lstlisting}[numbers=none]
livelit $dataframe at Dataframe { ... }
livelit $grade_cutoffs(avgs: List(Float))
  at (.A Float, .B Float, .C Float, .D Float) { ... }
\end{lstlisting}
The expansion type of \li{\$dataframe} is \li{Dataframe}
(a type classifying tabular floating point data together with string row and column keys, consistent with the expansion above)
and the expansion type of \li{\$grade_cutoffs} is a labeled product of grade cutoffs (Hazel notates labels \li{.label}).
Hazel summarizes the typing information from the livelit definition when the cursor is on the livelit's name,
just as it displays a function's type when its name is under the cursor (not shown).

@@

\paragraph{Parameters}\label{sec:parameterization} Parameters appear immediately after the name of the applied livelit.
For example, \li{averages} is provided as a parameter to \li{\$grade_cutoffs} in Fig.~\ref{fig:grading}.
The livelit declares a fixed number of parameters and specifies their types in its declaration.
For example, the declaration of \li{\$grade_cutoffs} shown above specifies one parameter of type \li{List(Float)}
(the grades data to be plotted, see below).

@@

To avoid this situation, the livelit expansion mechanism
maintains parameter and splice hygiene: parameters and splices are included in the expansion
in a capture-avoiding manner.
Consequently, clients can reason abstractly about binding in splices: variables in splices
always refer to bindings visible to the client, rather than bindings that are hidden inside expansions.
% This is implemented by alpha-renaming bindings internal to the expansion as necessary.
% (We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}.)

\subsubsection{Live Evaluation}\label{sec:live-evaluation}
Livelits can request the result of evaluating a splice or a parameter.
For example, the \li{\$dataframe} livelit uses this facility to display
the evaluation result for each cell when the cursor is not on it, like a spreadsheet.
The \li{\$grade_cutoffs} livelit uses this facility to plot the grades list, provided
as a parameter, on the number line.

@@

Similarly, the closure for the \li{\$grade_cutoffs} livelit in Fig.~\ref{fig:grading} includes a result for
the \li{grades} variable. However, this variable's result, along with the
results for other variables that depend on \li{grades}, such as \li{averages}, ultimately depend on the
expansion of the \li{\$dataframe} livelit. If we stop after the first phase of closure collection,
where these expansions have not yet been included,
then no useful result will be available here.
For this reason, there is a second phase of closure collection where any livelit holes that appear
in the collected livelit closures are \emph{resumed}, i.e. the expansion is generated
(in this case, for the \li{\$dataframe} livelit) and is used to fill
the hole before evaluation resumes, following the semantics for resumption developed by \citet{HazelnutLive}.
Livelit expansions do not themselves contain livelits, so no subsequent resumption phases are necessary.

@@

\paragraph{Indeterminate Results}
We used the phrase \emph{evaluation result} rather than \emph{value} above purposefully:
splices and parameters can be incomplete, i.e. they can contain holes, so
evaluation does not always result in a value \cite{HazelnutLive}.
In particular, when a hole appears in elimination position, e.g. in function position of a function application,
 the result is instead an indeterminate expression.
When a livelit requests an evaluation result, it must be able to handle indeterminate results.
For example, if the parameter to \li{\$grade_cutoffs} were a hole or some other indeterminate expression,
then it would have degraded functionality
(in this case, it would display the list elements that are values on the timeline, skipping indeterminate elements).
We will return to how livelit implementations handle this situation in Sec.~\ref{sec:livelit-definitions}.

>>>>

\section{Case Studies}\label{sec:case-studies}
Let us begin by considering livelits from the client programmer's perspective by way of
two domain-specific case studies:
a course grade assignment case study in Sec.~\ref{sec:live-grading}
and an image transformation case study in Sec.~\ref{sec:image-transformation}.
We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}.
These case studies and examples have been implemented\todo{make this not a lie}{}
in Hazel, a browser-based live programming environment for a pure typed functional language in the
ML family. We assume basic familiarity with ML.\todo{cite Elm tutorial for readers who are not familiar?}{}

\subsection{Case Study: Grading with Livelits}\label{sec:live-grading}
\begin{figure}
    \includegraphics[width=30pc]{grade-cutoff-livelit.png}
\caption{Grading Example}
\label{fig:grading}
\end{figure}

@@

Perhaps the most common end-user application in this domain is the spreadsheet.
A spreadsheet allows the instructor to record student grades using a natural tabular interface,
visualize this data in one of a finite number of styles, and perform basic computations on this data,
with results updated live, i.e. automatically after each cell has been edited.
However, these affordances are limited. It is difficult to package up common operations and workflows
into reusable libraries, interact with the data using domain-specific visualizations,
and perform operations that have not been anticipated by the designers of the spreadsheet system
(e.g. operations that prepare the data in an idiosyncratic format suitable for automated submission
to the university's grading system).
General-purpose programming languages equipped with suitable libraries for working with numeric data
are flexible enough to handle these scenarios, but users of these languages
sacrifice the ability to receive live feedback and directly manipulate the data and visualizations thereof.

Livelits are able to address this tension.
Fig.~\ref{fig:grading} shows a Hazel program where the user has:\todo{label each of these 1-5 in figure?}{}
\begin{enumerate}
\item recorded numeric \li{grades} for each student using a livelit, \li{\$dataframe}, that implements a tabular user interface with support for live evaluation of the Hazel expression in each cell,
\item programmatically computed the overall \li{averages} for each student by applying a function, \li{compute_averages}, defined in a library  (not shown) shared between multiple courses,
\item determined reasonable \li{cutoffs} between letter grades by directly manipulating a livelit that offers a domain-specific drag-and-drop interface superimposed on a live visualization of the grade distribution
(for the sake of space, we plot dots rather than a full histogram)\todo{should we do a histogram?}{},
\item programmatically assigned letter grades to each student based on the \li{cutoffs} determined in step (3) by calling another shared library function, \li{assign_grades},
\item formatted the final grade assignments in a manner suitable for submission to the university system by calling another shared library function, \li{format_for_university}.
\end{enumerate}

\subsubsection{Livelit Expansion}\label{sec:livelit-expansion}
Livelit expressions are given dynamic meaning by expansion to core expressions, i.e. expressions that do not contain livelits.
For example, the expansion of Fig.~\ref{fig:grading} is:\todo{change code once figure is final}{}

\begin{lstlisting}
let grades = Dataframe (
  ["A1", "A2", "A3", "A4", "Midterm", "Final"],
  [
    ("Cyrus", [0.0, 0.0, 0.0, 0.0, 0.0]),
    ("Nick", [0.0, 0.0, 0.0, 0.0, 0.0]),
    ("David", [0.0, 0.0, 0.0, 0.0, 0.0])
  ]) in
let averages = compute_averages(grades) in
let cutoffs = (.A 90.0, .B 80.0, .C 70.0, .D 60.0) in
format_for_university(assign_grades(averages, cutoffs))
\end{lstlisting}

The programmer can inspect this expansion in Hazel.
Ideally, however, it would not be necessary to inspect the expansion
(or the livelit implementation, which specifies the expansion logic and is discussed in Sec.~\ref{sec:livelit-definitions})
to reason about types and binding.
After all, one does not need to look inside
function bodies to reason about types and binding.
Instead, in the words of Reynolds\todo{cite}{},
``type structure is a syntactic discipline for maintaining levels of abstraction''.
The livelits mechanism supports reasoning abstractly about types and binding by
several means.

To support abstract reasoning about the type of the expansion,
livelit definitions declare an \emph{expansion type}.
The declarations of the two livelits in Fig.~\ref{fig:grading},
eliding their implementations, are:
\begin{lstlisting}[numbers=none]
livelit $dataframe at Dataframe { ... }
livelit $cutoffs(avgs: List(Float))
  at (.A Float, .B Float, .C Float, .D Float) { ... }
\end{lstlisting}
The expansion type of \li{\$dataframe} is \li{Dataframe}
(a type classifying tabular floating point data together with string row and column keys, consistent with the expansion above)
and the expansion type of \li{\$cutoffs} is a labeled product of grade cutoffs (Hazel notates labels \li{.label}).
Hazel summarizes the typing information from the livelit definition for the programmer when the cursor is on the livelit's name,
just as it displays a function's type when its name is under the cursor (not shown).\todo{could show if Hannah et al are finished}{}

@@

\paragraph{Parameters}\label{sec:parameterization} Parameters appear immediately after the name of the applied livelit.
For example, \li{averages} is provided as a parameter to \li{\$cutoffs} in Fig.~\ref{fig:grading}.
The livelit declares a fixed number of parameters and specifies their types in its declaration.
For example, the declaration of \li{\$cutoffs} shown above specifies one parameter of type \li{List(Float)}
(the grades data to be plotted, see below).

@@

To avoid this situation, the livelit expansion mechanism
maintains parameter and splice hygiene: parameters and splices are included in the expansion
in a capture-avoiding manner.
Consequently, clients can reason abstractly about binding in splices: variables in splices
always refer to bindings visible to the client, rather than bindings that are hidden inside expansions.
This is implemented by alpha-renaming bindings internal to the expansion as necessary.
(We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}.)

\subsubsection{Live Evaluation}\label{sec:live-evaluation}
Livelits can request the result of evaluating a splice or a parameter.
For example, the \li{\$dataframe} livelit uses this facility to display
the evaluation result for each cell when the cursor is not on it, like a spreadsheet.
The \li{\$cutoffs} livelit uses this facility to plot the grades list, provided
as a parameter, on the number line.

@@

Similarly, the closure for the \li{\$cutoffs} livelit in Fig.~\ref{fig:grading} includes a result for
the \li{grades} variable. However, this variable's result, along with the
results for other variables that depend on \li{grades}, such as \li{averages}, ultimately depend on the
expansion of the \li{\$dataframe} livelit. If we stop after the first phase of closure collection,
where these expansions have not yet been included,
then no useful result will be available here.
For this reason, there is a second phase of closure collection where any livelit holes that appear
in the collected livelit closures are \emph{resumed}, i.e. the expansion is generated
(in this case, for the \li{\$dataframe} livelit) and is used to fill
the hole before evaluation resumes, following the semantics for resumption developed by \citet{HazelnutLive}.
Livelit expansions do not themselves contain livelits, so no subsequent resumption phases are necessary.

@@

\paragraph{Indeterminate Results}
We used the phrase \emph{evaluation result} rather than \emph{value} above purposefully:
splices and parameters can be incomplete, i.e. they can contain holes, so
evaluation does not always result in a value \cite{HazelnutLive}.
In particular, when a hole appears in elimination position, e.g. in function position of a function application,
 the result is instead an indeterminate expression.
When a livelit requests an evaluation result, it must be able to handle indeterminate results.
For example, if the parameter to \li{\$cutoffs} were a hole or some other indeterminate expression\todo{small screenshot?}{},
then it would have degraded functionality
(in this case, it would display the list elements that are values on the timeline, skipping indeterminate elements\todo{implement this}{}).
We will return to how livelit implementations handle this situation in Sec.~\ref{sec:implementation}.
\todo{make sure we cover this}{}

### 2020-09-06 discussion.tex
popl21 reviews

\section{Discussion and Conclusion}\label{sec:discussion}\label{sec:conclusion}
Future work:
\begin{itemize}
  \item pattern matching
  \item type splices
  \item module splices
  \item explicitly make bindings available in splices
  \item better UI for closure provenance / connect with control flow / call stack better
  \item deriving livelits from type definitions
  \item bidirectional evaluation ala SnS
  \item integration with structure editing more cleanly
  \item \dots
  \item collaborative editing?
  \item side effects?
  \item full screening livelits (or compositions thereof) as a way to create end-user workflows
  \item big vision: authoring environment based on livelits
\end{itemize}

>>>>

\section{Discussion and Conclusion}\label{sec:discussion}\label{sec:conclusion}
\begin{quote}
  %
  \textit{
  %
  ``The arithmetical symbols are written diagrams and the geometrical figures are graphic formulas.''
  %
  }

  \vspace{3pt}

  \hfill{}--- David Hilbert~\cite{XXX}
  \end{quote}

Future work:
Recent approaches to deriving type-specific structure editors automatically from pretty printing logic \cite{hempeltiny} could perhaps 
be adapted to generate livelit implementations. Mechanisms for deriving simple 
livelit definitions from type definitions, perhaps similar to Haskell's \li{deriving} directive, 
may also prove helpful \cite{magalhaes2010generic}. The GEC toolkit has explored  
the automatic generation of simple user interfaces from type definitions \cite{DBLP:conf/afp/AchtenEPW04}.

\begin{itemize}
  \item pattern matching
  \item type splices
  \item module splices
  \item explicitly make bindings available in splices
  \item better UI for closure provenance / connect with control flow / call stack better
  \item deriving livelits from type definitions
  \item bidirectional evaluation ala SnS
  \item integration with structure editing more cleanly
  \item \dots
  \item collaborative editing?
  \item side effects?
  \item full screening livelits (or compositions thereof) as a way to create end-user workflows
  \item big vision: authoring environment based on livelits
\end{itemize}

### 2020-09-06 livelit-definitions.tex
popl21 reviews

Line 3 of Fig.~\ref{fig:color-impl} specifies the livelit's name, \li{\$color}, and its \emph{expansion type},
\li{Color}, which was discussed in Sec.~\ref{sec:livelit-expansion}.
The remainder of the definition is the livelit's implementation.
Livelits are implemented using a variation on the functional model-view-update
framework popularized by the Elm programming language \cite{ElmArchitecture}. We add a fourth component,
expansion generation. In addition, we add a simple monadic framework to provide the necessary  
interface between the livelit and the programming environment, discussed below, while retaining
a pure functional programming style.

\subsection{Model}\label{sec:model}
Line 6 of Fig.~\ref{fig:color-impl} specifies the livelit's \emph{model type},
here a labeled 4-tuple of \emph{splice references}, one for each of the four splices
that appear in the GUI in Fig.~\ref{fig:color}.
Models are persisted in the syntax tree of the program, so in practice,
 the system would check that the model type supports automatic serialization. 
 (Hazel is able to serialize all types because it is implemented as a 
 simple interpreter.)

The initial value of the model is specified starting on Line 7, 
which defines a command in the \li{UpdateMonad} that returns the 
initial model value after generating four new splice references
using the \li{new_splice} command:
\begin{lstlisting}[numbers=none]
new_splice : (Typ, Option(Exp)) -> UpdateMonad(SpliceRef)
\end{lstlisting}
This command creates a fresh splice
of the given type and, optionally, with the given initial expression,
and returns a unique splice reference, which uniquely and abstractly identifies that splice.
Bolded types like \li{Typ}, \li{Exp}, \li{UpdateMonad}, and \li{SpliceRef}
are provided by the standard library. The \li{Typ} and \li{Exp} types 
encode the syntax of Hazel's types and expressions.
These two arguments are provided here in quasiquoted form, e.g. \li{`0`} \cite{bawden1999quasiquotation}. 
If no initial expression is provided, the splice is initialized with 
an empty hole.

The system checks that the provided type and initial content is valid 
assuming only the explicitly specified set of captured bindings on Line 4.
Here, we do not need to capture any bindings because \li{Int} is a built in type
and the initial content is closed. However, 
for splices of user-defined types, those types need to be explicitly captured.
Similarly, helper functions that are used in initial splices (and in the expansion,
discussed below) must be explicitly captured. 
This serves to ensure \emph{context independence} -- the livelit needs to make no 
assumptions about the typing context at use sites.
We use an explicit capture set, rather than implicitly capture of all bindings at the definition site, 
both to prevent inadvertent capture 
and to ensure that the capture set is part
of the interface of the livelit definition, which simplifies matters related to 
module export and determining paths to the captured bindings at use sites \cite{TLMs}.
Formally, captured variables can be understood as parameters that have been 
immediately partially applied.

\subsection{Action}
Line 14 defines the \li{Action} type for the \li{\$color} livelit, which 
specifies a single user-initiated action: clicking on a color using the user 
interface on the right side of Fig.~\ref{fig:color}. Actions are emitted
from event handlers (e.g. click handlers) defined in the computed \li{view}, 
and actions are consumed by the \li{update} function, causing a change in the model. 
Let us discuss each of these functions in turn.

@@

When the model is updated, a new view is 
computed. The system then performs a diff between the old and new view in order to 
efficiently perform the necessary imperative updates to the editor's visual state.
Changes to splices can also cause the view to be recomputed, because the view might 
need to evaluate the splices. The \li{update} function does not itself 
have the ability to request evaluation, because the model should not depend directly  
which closure the user has selected. Of course, the \li{view} might emit 
result-dependent actions when appropriate.

>>>>

Line 1 of Fig.~\ref{fig:color-impl} specifies the livelit's name, \li{\$color}, and its \emph{expansion type},
\li{Color}, which was discussed in Sec.~\ref{sec:livelit-expansion}.
The remainder of the definition is the livelit's implementation.
Livelits are implemented using a variation on the functional model-view-update
framework popularized by the Elm programming language\todo{cites}. We add a fourth component,
expansion generation. In addition, we add a simple monadic framework to provide the necessary  
interface between the livelit and the programming environment, discussed below, while retaining
a pure functional programming model.\todo{cite monads in Haskell}

\subsection{Model}\label{sec:model}
\todo{final check of line numbers}{}Line 2 of Fig.~\ref{fig:color-impl} specifies the livelit's \emph{model type},
here a labeled 4-tuple of \emph{splice references}, one for each of the four splices
that appear in the GUI in Fig.~\ref{fig:color}.
Models are persisted in the syntax tree of the program, so in practice,
 the system would check that the model type supports automatic serialization. 
 (Hazel is able to serialize all types because it is implemented as a 
 simple interpreter.)

The initial value of the model is specified starting on Line 4, 
which defines a command in the \li{UpdateMonad} that returns the 
initial model value after generating four new splice references
using the \li{new_splice} command:
\begin{lstlisting}[numbers=none]
  new_splice : (Typ, Option(Exp)) -> UpdateMonad(SpliceRef)
\end{lstlisting}
This command creates a fresh splice
of the given type and, optionally, with the given initial expression,
and returns a unique splice reference, which uniquely and abstractly identifies that splice.
Bolded types like \li{Typ}, \li{Exp}, \li{UpdateMonad}, and \li{SpliceRef}
are provided by the standard library. The \li{Typ} and \li{Exp} types 
encode the syntax of Hazel's types and expressions.
These two arguments are provided here in quasiquoted form, e.g. \li{`0`}\todo{cite something}. 
If no initial expression is provided, the splice is initialized with 
an empty hole.

The system checks that the provided type and initial content is valid 
assuming only the explicitly specified set of captured bindings on Line 3.
Here, we do not need to capture anything because \li{Int} is a built in type
and the initial content is closed. However, 
for splices of user-defined types, those types need to be explicitly captured.
Similarly, helper functions that are used in initial splices (and in the expansion,
discussed below) must be explicitly captured. 
This serves to ensure \emph{context independence} -- the livelit needs to make no 
assumptions about the typing context at use sites.
We use an explicit capture set, rather than implicitly capture of all bindings at the definition site, 
both to prevent inadvertent capture\todo{cite heathers paper}{} 
and to ensure that the capture set is part
of the interface of the livelit definition, which simplifies matters related to 
module export and determining paths to the captured bindings at use sites \cite{TLMs}.

\subsection{Action}
Line 11 defines the \li{Action} type for the \li{\$color} livelit, which 
specifies a single user-initiated action: clicking on a color using the user 
interface on the right side of Fig.~\ref{fig:color}. Actions are emitted
from event handlers (e.g. click handlers) defined in the computed \li{view}, 
and actions are consumed by the \li{update} function, causing a change in the model. 
Let us discuss each of these functions in turn.

@@

When the model is updated, a new view is 
computed. The system then performs a diff between the old and new view in order to 
efficiently perform the necessary imperative updates to the editor's visual state.
Changes to splices can also cause the view to be recomputed, because the view might 
need to evaluate the splices\todo{optimizaiton?}. The \li{update} function does not itself 
have the ability to request evaluation, because the model should not depend directly  
which closure the user has selected. Of course, the \li{view} might emit 
result-dependent actions when appropriate.

### 2020-09-06 related-work.tex
popl21 reviews

\section{Related Work}\label{sec:related-work}
Cyrus has a list somewhere, ask if you want to work on this section

>>>>

\section{Related Work}\label{sec:related-work}
As detailed in Sec.~\ref{sec:intro},
the \emph{Graphite} system introduced the idea of filling 
typed holes using a type-specific user interface \cite{Graphite}.
This prior work engaged in substantial qualitative evaluation, 
which due to the fundamental similarities between the two systems 
is as relevant to our design as theirs. However, the prior work 
left a number of core technical issues unresolved, 
as summarized in Sec.~\ref{sec:intro}. 
Livelits resolve these issues in large part by bringing together 
seemingly disparate ideas from several other pieces of recent work.

In particular, recent work on type-specific languages (TSLs) \cite{TSLs} 
and typed literal macros (TLMs) \cite{TLMs}
explored similar ideas of user-defined
literal forms with support for hygienic splicing, with the former 
using type-directed dispatch similar to Graphite and the latter supporting 
decentralized extensibility via explicit naming as in our approach. 
However, these systems operate in a purely 
textual setting. There is no support for graphical notation or 
interaction, nor live feedback, which are the primary contributions of this paper.
The typed expansion judgement central to the typed livelit calculus in this paper 
is structured similarly 
to the corresponding judgements in the formal systems describing TLMs and TSLs,
and the reasoning principles (e.g. hygiene properties, the focus on reasoning abstractly 
about typing) was inspired directly by this work. However, the  
approach to hygiene we take is cleaner by its use of function application 
rather than direct insertion of splices (and could perhaps be ported to those formalisms).
Splices also operate quite differently in our work,
because they are placed automatically and structurally delimited in the user interface, rather than 
placed by the client and then parsed out of the text by a custom parser charged 
with retaining provenance information.
This substantially complicates the design of splices in the prior work.

This structural delimitation of different notations is reminiscent of work on 
\emph{language boxes} \cite{DBLP:conf/sle/RenggliDN09}, which focused primarily on 
combining different notations, primarily textual, using structural delimiters inserted 
explicitly by the user using a special-purpose editor.,

There has been a long line of research on \emph{projectional editing}, where
the user edits graphical representations (projections) of code constructs \cite{DBLP:journals/ile/MillerPMV94,read1996generating,DBLP:conf/sde/Reiss84,DBLP:conf/uist/KoM05}. 
Many of the oldest systems offer only a fixed set of projections and interaction techniques. 
More recently, language workbenches with support for projectional editing like Citrus \cite{DBLP:conf/uist/KoM05}
and MPS \cite{voelter2011language} have made it easier to define new projectional editors.
However, these systems generate entire editors, whereas livelit definitions can be 
included granularly in a livelit context and support decentralized extensibility.
Furthermore, livelit definitions are governed by a type and binding discipline that 
serves to ensure that a programmer unfamiliar with the semantics of an unfamiliar livelit
can still be sure that it is not operating outside of the bounds of the type system.
In many ways, an unfamiliar livelit can be approached much like an unfamiliar function
can be approached: by inspecting its type signature and reasoning compositionally 
about arguments, i.e. splices.
Finally, livelits integrate cleanly with live evaluation of programs with holes \cite{HazelnutLive},
whereas other projectional editors either do not support evaluation of holes or do 
not use that information to provide feedback to the user.

Related to projectional editors are ``syntax-recognizing editors'' \cite{DBLP:journals/tosem/BallanceGV92,DBLP:conf/sde/HorganM84} and 
a great variety of systems that generate visualizations from code \cite{DBLP:conf/chi/Lerner20,koschke2003software,urquiza2004survey}. Livelits differ 
from these systems in directionality: visualization systems are given code and generate
a visualization, whereas livelits are asked to create code where there would otherwise
be a hole based on interactions that the programmer has with a visual interface.

Conal Elliott's work on tangible functional programming \cite{DBLP:conf/icfp/Elliott07} similarly explored a
system that provides visual feedback about the result of a computation, 
but the computation itself could be constructed using only a fixed structural syntax
manifest as a series of connected windows.

The Vital programming environment for Haskell \cite{hanna2002interactive} supports type-specific stylesheets
that can be used to create custom user interfaces for both editing and displaying
values. The editors support splices, called \emph{cells}, that can contain 
Haskell code. Results are computed in a live manner, though there is no support
for evaluating incomplete programs. The visualizations themselves also cannot provide
live feedback. Moreover, the system does not enforce any hygiene or typing principles:
the user is entirely responsible for syntactic and semantic correctness, and the 
underlying semantics have not been formally defined.

% Hazelnut
% Hazelnut Live
% Its Alive! paper (GUI widgets)
% Tangible FUnctional Programming 
% Vital https://dl.acm.org/doi/10.1145/581478.581493 and https://link.springer.com/chapter/10.1007/11964681_12
% Projection Boxes http://cseweb.ucsd.edu/~lerner/proj-boxes.html
% Tiny Structure Editors for Low Low Prices
% Typed Holes Work (synthesis stuff from Haskell, mention Haskell)
% Pharo / Smalltalk 
% GEC Toolkit: https://link.springer.com/chapter/10.1007/11546382_5
% Previous use of live literals https://homepages.cwi.nl/~storm/livelit/livelit.html
% Visual Haskell and related visual programming environments: https://pdfs.semanticscholar.org/6e43/5494ef1aaaa6ec0fcf47bb5b03d7c06ef937.pdf
% widgets in dataflow languages like puredata and labview
